Ordinal Classification Using Partially Ordered Feature Sets\tnoteref{t1}

### 2024-11-15 paper/main.tex
Add files via upload

@@

\usepackage{a4wide}
\usepackage{amsmath, amsfonts, amssymb}
\usepackage{pb-diagram}
\usepackage{graphicx}
\usepackage{subfigure}

@@

\author[mipt]{M.\,P.~Kuznetsov\corref{cor1}}
\ead{mikhail.kuznecov@phystech.edu}
\author[mipt,ccas]{V.\,V.~Strijov}
\ead{strijov@ccas.ru}
\cortext[cor1]{Corresponding author}
\address[mipt]{Moscow Institute of Physics and Technology, Institutskiy lane 9, Dolgoprudny city, Moscow region, 141700, Russia}
\address[ccas]{Dorodnicyn Computing Center of Russian Academy of Sciences,  Vavilov st. 40, 119333 Moscow, Russia}

@@

\section{Introduction}
We consider an ordinal classification problem using partially ordered feature sets. This problem generalizes the well-known ordinal classification problem with monotonic constraints~\cite{Kotlowski2013, Corrente2013}. The objective is to construct a monotonic function, further calling a classification model, mapping an object set to a class label set such that the class label set is strictly totally ordered. The requirement of the classification model monotonicity is a form of prior knowledge of a relation between a response and predictor variables~\cite{raey}. The motivation for this type of problem statement arises in areas of support decision theory, information retrieval~\cite{Schafer2007,Trotman2005,Spirin2011} and preference learning~\cite{Fuernkranz2011}.

@@

The proposed method is compared with decision tree and generalized linear regression methods on synthetic and real data. As a real data we consider a problem of the IUCN Red List categorization, also described in. The goal is to categorize those plants and animals that are facing the higher risk of extinction. The following categories are considered: least concern, near threatened, vulnerable, endangered, critically endangered, extinct in the wild. This categorization is monotonic. The partially ordered features (e.g., population size, area square) are given by the experts. The proposed method shows better results in comparison with the alternative approaches.

@@

Let us define a partial order matrix~$\bZ_j$ for the sample~$\mathfrak{D}$ and every set~$X_j$ such that the matrix describes binary relation between each pair of the sample elements,
\[
\bZ_j(i,k) = z_j(\bx_i,\bx_k),
\]
where~$j$~ is a feature index, and~$i,k$~are the object indices.
Note that for the matrix~$\bZ_j$, as well as for the function~$z_j$, elements~$\bZ_j(i,k)$~and~$\bZ_j(k,i)$ do not depend on each other. For example, the objects~$i,k$ can be incomparable with each other for the feature~$j$. In this case for the matrix~$\bZ_j$ the following equalities hold:
\[
\bZ_j(i,k)=\bZ_j(k,i)=0.
\]
On the other hand, the objects~$i,k$ can be equal for the feature~$j$. In this case, the following equalities hold:
\[
\bZ_j(i,k)=\bZ_j(k,i)=1.
\]
\begin{Ex}
The partially ordered set~$X_j$ consists of four objects with the following binary relation,
\[
X_j=\{x_{1},x_{2},x_{3},x_{4}|\;x_{1}\succeq x_{2}, x_{2}\succeq x_{3}, x_{2}\succeq x_{4}\}.
\]
The partial order on the objects~$x_{1},x_{2},x_{3},x_{4}$ is described by a partial order graph,
\[
\begin{diagram}
\node{x_{1}} \arrow[2]{e}{}
\node[2]{x_{2}} \arrow[2]{}
\arrow{se,b}{}
\node[2]{x_{3}} \\
\node[4]{x_{4}}
\end{diagram}
\]
This graph corresponds to a matrix~$\bZ_j$:
\[
\bZ_j=
\begin{pmatrix}
1 & 1 & 1 & 1 \\
0 & 1 & 1 & 1 \\
0 & 0 & 1 & 0 \\
0 & 0 & 0 & 1 \\
\end{pmatrix}.
\]
\label{ex1}
\end{Ex}
The class label set
\begin{equation}
Y=\{l_1,l_2,...,l_K\},\quad y_i\in Y
\label{yValues}
\end{equation}
is a finite set with the specified strict linear order relation,~$l_1\prec l_2,...,\prec l_K$, where~$K$~is a number of the class labels. Similarly to the matrices~$\bZ_1,...,\bZ_n$, construct a partial order matrix~$\bZ_0$ corresponding to the vector of class labels~$\by=[y_1,...,y_m]\T$ of the sample~$\mathfrak{D}$:
\[
\bZ_0(i,k) =
\begin{cases}
1,\quad\text{if } y_i\succeq y_k, \\
0,\quad\text{otherwise}.
\end{cases}
\]
\paragraph{Partial order cone}
Introduce a polyhedral cone~$\mathcal{X}$, corresponding to the partially ordered set of feature values.
\begin{Def}
A polyhedral cone in~$\mathbb{R}^m$~is a set~$\mathcal{X}$ such that
\[
\mathcal{X}=\{\bchi\;|\;\bA\bchi\leq \mathbf{0},\;\bchi\in\mathbb{R}^m\},
\]
fore some matrix~$\bA$.
\end{Def}
For each set~$X_j$ define a cone~$\mathcal{X}_j$ in a space~$\mathbb{R}^m_+$:
\begin{equation}
\mathcal{X}_j=\{\bchi_j\in\mathbb{R}^m_+|\quad x_{ij}\succeq x_{kj}\rightarrow \chi_{ij}\geq \chi_{kj}\quad \forall i,k=1,...,m\},
\label{ConeDef}
\end{equation}
where~$x_{ij}$~and~$x_{kj}$~are values of the feature~$j$ on the sample elements~$i$~and~$k$, respectively.
For the cone~$\mathcal{X}$ (index~$j$ is omitted for the reasons of convenience) the following theorem holds.
\begin{Theorem}
A vector~$\bchi$ belonging to the cone~$\mathcal{X}$ can be represented as the nonnegative combination of cone generators,
$$\bchi=\sum\limits_{k=1}^{m}\lambda_{k}\bzeta_{k}, \quad\lambda_{k}\geq 0,$$
where~$\bzeta_k$~is a generatrix of the cone~$\mathcal{X}$,
\[
\bzeta_{k}(i)=
\begin{cases}
1,\quad \text{if}\quad x_i\succeq x_k,\\
0,\quad\text{if}\quad x_i\nsucceq x_k,
\end{cases}
\]
and this decomposition is unique.
\label{thConeGen}
\end{Theorem}
\begin{figure}[h]
\begin{center}
\includegraphics[width=.6\textwidth]{coneTheorem.eps}
\caption{Vector decomposition on the cone generators}
\label{fig:ConeTheorem}
\end{center}
\end{figure}
The fig.~\ref{fig:ConeTheorem} illustrates vector~$\bchi$ decomposition on the generators~$\bzeta_1,\bzeta_2$ of the cone~$\mathcal{X}$. The vector~$\bchi\in \mathcal{X}$ is a nonnegative combination of the generators~$\bzeta_1=[1;1]$~and~$\bzeta_2=[1;0]$ with the coefficients~$\lambda_1,\lambda_2\geq 0$.

@@

\paragraph{Loss function minimization for the optimal parameters search}
To solve the problem of ordinal classification with monotonic constraints~(\ref{ErrGen}), let us to formulate a problem of finding the optimal parameters of the model~(\ref{utilityDecompDual}).
According to formula~(\ref{ErrGen}), the optimal parameters~$\hat{\bw},\hat{\blambda}$ minimize the loss function
\[
(\hat{\bw},\hat{\blambda})=\arg\min\limits_{\bw,\blambda}\left(S(\mathfrak{D}|\bw,\blambda)\right)=\arg\min\limits_{\bw,\blambda}\left(\sum\limits_{i=1}^m{s(f_{\bw,\blambda}(\bx_i), y_i)}\right),
\]
where~$s(f_{\bw,\blambda}(\bx_i), y_i)$~--- is the value of loss on an object~$\bx_i$, and a monotonic function
\begin{equation}
f_{\bw,\blambda}(\bx_i) = \phi\left(\sum\limits_{k=1}^m \lambda_{k} \Psi(\bx_i, \bx_k)\right)
\label{ordClassMdl}
\end{equation}
is defined by the expressions~(\ref{problOrdClass})~and~(\ref{utilityDecompDual}).
The explicit form of the loss function~$S$, of the decision rule~$\phi$ and of the binary relation function~$\Psi$ will be considered in the next section.
\section{An ordinal classification algorithm}
To estimate parameters of the model~(\ref{ordClassMdl}) we propose a two-stage algorithm. In the first stage estimate the partial order matrix~$\bPsi$, where the elements of the matrix~$\bPsi$ equal to the values of function~$\Psi$ on the objects of the sample~$\mathfrak{D}$,
\[
\bPsi(i,k) = \Psi(\bx_i,\bx_k).
\]
In the second stage estimate parameters~$\lambda_k$ of a linear combination~$\sum\limits_{k=1}^m \lambda_{k} \Psi(\bx_i, \bx_k)$ from~(\ref{utilityDecompDual}).
\paragraph{Partial order matrix estimation}
Estimate matrix~$\bPsi$ of the pairwise objects dominance using matrices~$\bZ_1,...,\bZ_n$ corresponding to the partially ordered feature sets. Let every element of a matrix~$\bPsi$ be a linear combination of elements of the matrices~$\bZ_1,...,\bZ_n$, according to~(\ref{problLinearMon}):
\[
\bPsi(i,k)=\sum\limits_{j=1}^n w_j\bZ_j(i,k).
\]
The optimal parameters~$\hat{\bw}$ minimize a loss function
\[
\hat{\bw} = \arg\min\limits_{\bw}\sum\limits_{i=1}^m\sum\limits_{k=1}^m\left(\bZ_0(i,k) - \sum\limits_{j=1}^n w_j\bZ_j(i,k)\right)^2,
\]
where~$\bZ_0$~is a partial order matrix corresponding to the class label vector~$\by$.
Since the target variable~$\bZ_0(i,k)$ is binary, we propose to estimate an optimal parameter vector~$\hat{\bw}$ using one of the standard two-class classification methods. In this paper we use the logistic regression method.
\begin{Ex}
Give an example of a linear combination of matrices~$\bZ_1,\bZ_2$ corresponding to the agreed partial orders. Consider a set consisting of the four elements~$X=\{x_1,x_2,x_3,x_4\}$ and the matrix~$\bZ_1$ from Example~\ref{ex1},
\[
\bZ_1=
\begin{pmatrix}
1 & 1 & 1 & 1 \\
0 & 1 & 1 & 1 \\
0 & 0 & 1 & 0 \\
0 & 0 & 0 & 1 \\
\end{pmatrix},
\]
corresponding to the partial order
\[
\begin{diagram}
\node{x_{1}} \arrow[2]{e}{}
\node[2]{x_{2}} \arrow[2]{}
\arrow{se,b}{}
\node[2]{x_{3}} \\
\node[4]{x_{4}} %\arrow{ne,r}{C}
\end{diagram}
\]
As a second matrix~$\bZ_2$, consider
\[
\bZ_2=
\begin{pmatrix}
1 & 1 & 1 & 1 \\
0 & 1 & 1 & 1 \\
0 & 0 & 1 & 1 \\
0 & 0 & 0 & 1 \\
\end{pmatrix},
\]
corresponding to the linear order
\[
\begin{diagram}
\node{x_{1}} \arrow[2]{e}
\node[2]{x_{2}} \arrow[2]{e}
\node[2]{x_{3}} \arrow[2]{e}
\node[2]{x_{4}}
\end{diagram}
\]
Construct a matrix~$\bPsi$ as a linear combination of the matrices~$\bZ_1,\bZ_2$ with the parameters~$w_1=w_2=\frac{1}{2}$:
\[
\bPsi=\frac{1}{2}\bZ_1+\frac{1}{2}\bZ_2=
\begin{pmatrix}
1 & 1 & 1 & 1 \\
0 & 1 & 1 & 1 \\
0 & 0 & 1 & 1/2 \\
0 & 0 & 0 & 1 \\
\end{pmatrix}.
\]
Since the matrices~$\bZ_1$~and~$\bZ_2$ correspond to the almost equivalent object orderings, the only element of the matrix~$\bPsi$ equals~$\frac{1}{2}$. This fact can be interpreted as uncertainty in the dominance of the only pair of objects~$x_3$~and~$x_4$.
\end{Ex}

>>>>

\usepackage{algorithm}
\usepackage{algorithmic}

@@

\usepackage{a4wide}
\usepackage{amsmath, amsfonts, amssymb}
\usepackage{pb-diagram}
\usepackage{graphicx}
\usepackage{subfigure}
\usepackage{tikz}

\usepackage[dvipsnames, svgnames, x11names]{xcolor}
\usepackage{color}

@@

\author[mipt]{I.\,D.~Papay}
\ead{papai.id@phystech.edu}
\author[mipt]{M.\,P.~Kuznetsov\corref{cor1}}
\ead{mikhail.kuznecov@phystech.edu}
\author[mipt,ccas]{V.\,V.~Strijov}
\ead{strijov@ccas.ru}
\cortext[cor1]{Corresponding author}
\address[mipt]{Moscow Institute of Physics and Technology, Institutskiy lane 9, Dolgoprudny city, Moscow region, 141700, Russia}
\address[ccas]{Dorodnicyn Computing Center of Russian Academy of Sciences,  Vavilov st. 40, 119333 Moscow, Russia}

@@

\section{Introduction}
We consider an ordinal classification problem using partially ordered feature sets. This problem generalizes the well-known ordinal classification problem with monotonic constraints~\cite{Kotlowski2013, Corrente2013, Ali2020}. The objective is to construct a monotonic function, further calling a classification model, mapping an object set to a class label set such that the class label set is strictly totally ordered. The requirement of the classification model monotonicity is a form of prior knowledge of a relation between a response and predictor variables~\cite{raey, Jorge2024, Eduardo2022}. The motivation for this type of problem statement arises in areas of support decision theory~\cite{Enrique2023,Ameed2022,Yossi2022}, information retrieval~\cite{Schafer2007,Trotman2005,Spirin2011} and preference learning~\cite{Fuernkranz2011, Bofei2024}.

The task is set as follows. There are a number of experts ranking some partially ordered set of objects. Our goal is to teach the model to predict the rank~\cite{Nadia2020} of an individual object for any other expert~\cite{Elena2023}. At the same time, all that is known about an object is a finite-dimensional vector of its features.

@@

The proposed method is compared with decision tree, graph models~\cite{Wilhelm2021} and generalized linear regression methods~\cite{Eugene2023} on synthetic and real data. As a real data we consider a problem of the IUCN Red List categorization, also described in. The goal is to categorize those plants and animals that are facing the higher risk of extinction. The following categories are considered: least concern, near threatened, vulnerable, endangered, critically endangered, extinct in the wild. This categorization is monotonic. The partially ordered features (e.g., population size, area square) are given by the experts. The proposed method shows better results in comparison with the alternative approaches.

@@

Let us define a partial order matrix~$\bZ_j$ for the sample~$\mathfrak{D}$ and every set~$X_j$ such that the matrix describes binary relation between each pair of the sample elements,
\[
\bZ_j(i,k) = z_j(\bx_i,\bx_k),
\]
where~$j$~ is a feature index, and~$i,k$~are the object indices.
Note that for the matrix~$\bZ_j$, as well as for the function~$z_j$, elements~$\bZ_j(i,k)$~and~$\bZ_j(k,i)$ do not depend on each other. For example, the objects~$i,k$ can be incomparable with each other for the feature~$j$. In this case for the matrix~$\bZ_j$ the following equalities hold:
\[
\bZ_j(i,k)=\bZ_j(k,i)=0.
\]
On the other hand, the objects~$i,k$ can be equal for the feature~$j$. In this case, the following equalities hold:
\[
\bZ_j(i,k)=\bZ_j(k,i)=1.
\]
\begin{Ex}
The partially ordered set~$X_j$ consists of four objects with the following binary relation,
\[
X_j=\{x_{1},x_{2},x_{3},x_{4}|\;x_{1}\succeq x_{2}, x_{2}\succeq x_{3}, x_{2}\succeq x_{4}\}.
\]
The partial order on the objects~$x_{1},x_{2},x_{3},x_{4}$ is described by a partial order graph,
\[
\begin{diagram}
\node{x_{1}} \arrow[2]{e}{}
\node[2]{x_{2}} \arrow[2]{}
\arrow{se,b}{}
\node[2]{x_{3}} \\
\node[4]{x_{4}}
\end{diagram}
\]
This graph corresponds to a matrix~$\bZ_j$:
\[
\bZ_j=
\begin{pmatrix}
1 & 1 & 1 & 1 \\
0 & 1 & 1 & 1 \\
0 & 0 & 1 & 0 \\
0 & 0 & 0 & 1 \\
\end{pmatrix}.
\]
\label{ex1}
\end{Ex}
The class label set
\begin{equation}
Y=\{l_1,l_2,...,l_K\},\quad y_i\in Y
\label{yValues}
\end{equation}
is a finite set with the specified strict linear order relation,~$l_1\prec l_2,...,\prec l_K$, where~$K$~is a number of the class labels. Similarly to the matrices~$\bZ_1,...,\bZ_n$, construct a partial order matrix~$\bZ_0$ corresponding to the vector of class labels~$\by=[y_1,...,y_m]\T$ of the sample~$\mathfrak{D}$:
\[
\bZ_0(i,k) =
\begin{cases}
1,\quad\text{if } y_i\succeq y_k, \\
0,\quad\text{otherwise}.
\end{cases}
\]
\paragraph{Partial order cone}
Introduce a polyhedral cone~$\mathcal{X}$, corresponding to the partially ordered set of feature values.
\begin{Def}
A polyhedral cone in~$\mathbb{R}^m$~is a set~$\mathcal{X}$ such that
\[
\mathcal{X}=\{\bchi\;|\;\bA\bchi\leq \mathbf{0},\;\bchi\in\mathbb{R}^m\},
\]
fore some matrix~$\bA$.
\end{Def}

By matrix $A$ in our article, from here on, we consider the matrix
\[
\bA_m=
\begin{pmatrix}
-1 & 1 & 0 & \dots & 0 & 0 \\
0 & -1 & 1 & \dots & 0 & 0 \\
\vdots & \vdots & \vdots & \ddots & \vdots & \vdots \\
0 & 0 & 0 & \dots & -1 & 1 \\
0 & 0 & 0 & \dots & 0 & -1 \\
\end{pmatrix},
\]

For each set~$X_j$ define a cone~$\mathcal{X}_j$ in a space~$\mathbb{R}^m_+$:
\begin{equation}
\mathcal{X}_j=\{\bchi_j\in\mathbb{R}^m_+|\quad x_{ij}\succeq x_{kj}\rightarrow \chi_{ij}\geq \chi_{kj}\quad \forall i,k=1,...,m\},
\label{ConeDef}
\end{equation}
where~$x_{ij}$~and~$x_{kj}$~are values of the feature~$j$ on the sample elements~$i$~and~$k$, respectively.
For the cone~$\mathcal{X}$ (index~$j$ is omitted for the reasons of convenience) the following theorem holds.
\begin{Theorem}
A vector~$\bchi$ belonging to the cone~$\mathcal{X}$ can be represented as the nonnegative combination of cone generators,
$$\bchi=\sum\limits_{k=1}^{m}\lambda_{k}\bzeta_{k}, \quad\lambda_{k}\geq 0,$$
where~$\bzeta_k$~is a generatrix of the cone~$\mathcal{X}$,
\[
\bzeta_{k}(i)=
\begin{cases}
1,\quad \text{if}\quad x_i\succeq x_k,\\
0,\quad\text{if}\quad x_i\nsucceq x_k,
\end{cases}
\]
and this decomposition is unique.

\begin{tikzpicture}[>=stealth]
\draw[thick, ->, black] (0,0) -- (12,0);
\draw[thick, ->, red] (0,0) -- (6,0);

\draw[thick, ->, black] (0,0) -- (0,12);

\draw[ultra thick, black] (0,0) -- (9,9);
\draw[ultra thick, ->, red] (0,0) -- (4.5,4.5);

\draw[thick, blue] (5.25,5.25) -- (8.25, 2.625) -- (8.25, 0);

\draw[dashed] (9,9) -- (11.25,0);

\node[below left] at (0,12) {\color{black}$\mathcal{X}_1$};
\node[below right] at (12,0) {\color{black}$\mathcal{X}_2$};
\node[above right] at (9,9) {\color{black}$\mathcal{X}$};
\node[above right] at (8.25, 2.625) {\color{black}$\mathcal{X}$};

\node[below right] at (5.85,0) {\color{red}$\bzeta_{2}$};
\node[above right] at (4.0,4.5) {\color{red}$\bzeta_{1}$};

\node[below right] at (8.45,1.3) {\color{blue}$\lambda_{2}$};
\node[above right] at (7.0,3.5) {\color{blue}$\lambda_{1}$};

\end{tikzpicture}

The figure illustrates vector~$\bchi$ decomposition on the generators~$\bzeta_1,\bzeta_2$ of the cone~$\mathcal{X}$. The vector~$\bchi\in \mathcal{X}$ is a nonnegative combination of the generators~$\bzeta_1=[1;1]$~and~$\bzeta_2=[1;0]$ with the coefficients~$\lambda_1,\lambda_2\geq 0$.

@@

\paragraph{Loss function minimization for the optimal parameters search}
To solve the problem of ordinal classification with monotonic constraints~(\ref{ErrGen}), let us to formulate a problem of finding the optimal parameters of the model~(\ref{utilityDecompDual}).
According to formula~(\ref{ErrGen}), the optimal parameters~$\hat{\bw},\hat{\blambda}$ minimize the loss function
\[
(\hat{\bw},\hat{\blambda})=\arg\min\limits_{\bw,\blambda}\left(S(\mathfrak{D}|\bw,\blambda)\right)=\arg\min\limits_{\bw,\blambda}\left(\sum\limits_{i=1}^m{s(f_{\bw,\blambda}(\bx_i), y_i)}\right),
\]
where~$s(f_{\bw,\blambda}(\bx_i), y_i)$~--- is the value of loss on an object~$\bx_i$, and a monotonic function
\begin{equation}
f_{\bw,\blambda}(\bx_i) = \phi\left(\sum\limits_{k=1}^m \lambda_{k} \Psi(\bx_i, \bx_k)\right)
\label{ordClassMdl}
\end{equation}
is defined by the expressions~(\ref{problOrdClass})~and~(\ref{utilityDecompDual}).
The explicit form of the loss function~$S$, of the decision rule~$\phi$ and of the binary relation function~$\Psi$ will be considered in the next section.
\section{Classification methods}
\subsection{Isothonic regression}

Isotonic regression is a technique used to fit a non-decreasing function to a set of data points.
We can think of isotonic regression as a generalization of linear regression where the function is constrained to be non-decreasing.

More formally suppose that one collects $n$ sample points $y_1,\dots, y_n$ and has weights $w_1, \dots, w_n$ associated with each data point.

When no a priori weights are given, we can set $w_i = 1$ for all $i \in \llbracket 1, n \rrbracket$.

Then, the isotonic regression problem can be formulated as follows:

\begin{align}
\min_{x \in \mathbb{R}^n}
\frac{1}{2}
& \sum_{i=1}^{n} w_i (x_i - y_i)^2 \\
\text{s.t.} & \quad x_1 \leq x_2 \leq \ldots \leq x_n \nonumber
\end{align}

The formulation of the isotonic regression problem given in (10) is a quadratic program (QP) with linear constraints.
Using a matrix notation, introducing the matrix
$$A = \begin{pmatrix} 1 & -1 &                       &                       &    \\
  & 1  & -1                    &                       &    \\
  &    & \ddots & \ddots &    \\
  &    &                       & 1                     & -1\end{pmatrix}
 =
\begin{pmatrix}
a_1^\top \\
\vdots   \\
a_{n-1}^\top
\end{pmatrix}
 \in \mathbb{R}^{n-1 \times n}\enspace,
$$
the isotonic regression problem can be formulated as follows:

\begin{align}
\min_{x \in \mathbb{R}^n}
\frac{1}{2}
& \sum_{i=1}^{n} w_i (x_i - y_i)^2 \\
\text{s.t.} & \quad Ax \leq 0 \nonumber
\end{align}

In what follows we will write $W=\diag(w_1,\dots,w_n)$, assuming that $w_i > 0$ for all $i \in \llbracket 1, n \rrbracket$, so $W^{-1}= \diag(1/w_1,\dots,1/w_n)$.
Hence, we can rewrite the isotonic regression problem as

\begin{align}
\min_{x \in \mathbb{R}^n}
\frac{1}{2}
& (x-y)^\top W (x-y) \\
\text{s.t.} & \quad Ax \leq 0 \enspace.
\end{align}

\begin{algorithm}[ht]
\caption{PAVA}\label{alg:cap}
\begin{algorithmic}
\Require $y \in\mathbb{R}^n, w \in \mathbb{R}^n_{+}$
\STATE $r \leftarrow y$
\STATE $W \leftarrow y$
\STATE $J = [\{1\},\dots, \{n\}]$ \COMMENT{lists of blocks}
\STATE $i=1$ (index of list start at 0 here)
\WHILE{$i<n$}
    \IF{$r_i < r_{i-1}$} \COMMENT{Find adjacent violators and merge groups}
        \State $r_i \leftarrow \frac{W_i r_i + W_{i-1} r_{i-1}}{W_i + W_{i-1}}$
        \State $W_i \leftarrow W_i + W_{i-1}$
        \State $J_i \leftarrow J_i \cup J_{i-1}$
        \State Remove $r_{i-1}$, $W_{i-1}$ and $J_{i-1}$ from the lists
        \IF{$i > 1$}
            \STATE{$i\leftarrow i-1$}
        \ENDIF
    \ELSE
        \STATE{$i  \leftarrow i + 1$}
    \ENDIF
\ENDWHILE
\FOR{$i=1$ to len($J$)}
    \STATE{$r_{J_i} \leftarrow \bar{y}_{J_i} \mathbf{1}_{J_i}$}
    \COMMENT{Set the block to the average value}
\ENDFOR
\RETURN{$r, J$}
\end{algorithmic}
\end{algorithm}

Hence, the isotonic regression problem can be formulated as a convex quadratic program with linear constraints.

The intuition behind the PAVA algorithm is to merge adjacent blocks of constant values in the primal vector, as the targeted solution is simply the average of the observe signal over each block.
Hence, the algorithm aims at creating the blocks of constant values in the primal vector. The dual variables can be inferred from the  solution from the previous lemma.

\subsection{An ordinal classification algorithm}
To estimate parameters of the model~(\ref{ordClassMdl}) we propose a two-stage algorithm. In the first stage estimate the partial order matrix~$\bPsi$, where the elements of the matrix~$\bPsi$ equal to the values of function~$\Psi$ on the objects of the sample~$\mathfrak{D}$,
\[
\bPsi(i,k) = \Psi(\bx_i,\bx_k).
\]
In the second stage estimate parameters~$\lambda_k$ of a linear combination~$\sum\limits_{k=1}^m \lambda_{k} \Psi(\bx_i, \bx_k)$ from~(\ref{utilityDecompDual}).
\paragraph{Partial order matrix estimation}
Estimate matrix~$\bPsi$ of the pairwise objects dominance using matrices~$\bZ_1,...,\bZ_n$ corresponding to the partially ordered feature sets. Let every element of a matrix~$\bPsi$ be a linear combination of elements of the matrices~$\bZ_1,...,\bZ_n$, according to~(\ref{problLinearMon}):
\[
\bPsi(i,k)=\sum\limits_{j=1}^n w_j\bZ_j(i,k).
\]
The optimal parameters~$\hat{\bw}$ minimize a loss function
\[
\hat{\bw} = \arg\min\limits_{\bw}\sum\limits_{i=1}^m\sum\limits_{k=1}^m\left(\bZ_0(i,k) - \sum\limits_{j=1}^n w_j\bZ_j(i,k)\right)^2,
\]
where~$\bZ_0$~is a partial order matrix corresponding to the class label vector~$\by$.
Since the target variable~$\bZ_0(i,k)$ is binary, we propose to estimate an optimal parameter vector~$\hat{\bw}$ using one of the standard two-class classification methods. In this paper we use the logistic regression method.
\begin{Ex}
Give an example of a linear combination of matrices~$\bZ_1,\bZ_2$ corresponding to the agreed partial orders. Consider a set consisting of the four elements~$X=\{x_1,x_2,x_3,x_4\}$ and the matrix~$\bZ_1$ from Example~\ref{ex1},
\[
\bZ_1=
\begin{pmatrix}
1 & 1 & 1 & 1 \\
0 & 1 & 1 & 1 \\
0 & 0 & 1 & 0 \\
0 & 0 & 0 & 1 \\
\end{pmatrix},
\]
corresponding to the partial order
\[
\begin{diagram}
\node{x_{1}} \arrow[2]{e}{}
\node[2]{x_{2}} \arrow[2]{}
\arrow{se,b}{}
\node[2]{x_{3}} \\
\node[4]{x_{4}} %\arrow{ne,r}{C}
\end{diagram}
\]
As a second matrix~$\bZ_2$, consider
\[
\bZ_2=
\begin{pmatrix}
1 & 1 & 1 & 1 \\
0 & 1 & 1 & 1 \\
0 & 0 & 1 & 1 \\
0 & 0 & 0 & 1 \\
\end{pmatrix},
\]
corresponding to the linear order
\[
\begin{diagram}
\node{x_{1}} \arrow[2]{e}
\node[2]{x_{2}} \arrow[2]{e}
\node[2]{x_{3}} \arrow[2]{e}
\node[2]{x_{4}}
\end{diagram}
\]
Construct a matrix~$\bPsi$ as a linear combination of the matrices~$\bZ_1,\bZ_2$ with the parameters~$w_1=w_2=\frac{1}{2}$:
\[
\bPsi=\frac{1}{2}\bZ_1+\frac{1}{2}\bZ_2=
\begin{pmatrix}
1 & 1 & 1 & 1 \\
0 & 1 & 1 & 1 \\
0 & 0 & 1 & 1/2 \\
0 & 0 & 0 & 1 \\
\end{pmatrix}.
\]
Since the matrices~$\bZ_1$~and~$\bZ_2$ correspond to the almost equivalent object orderings, the only element of the matrix~$\bPsi$ equals~$\frac{1}{2}$. This fact can be interpreted as uncertainty in the dominance of the only pair of objects~$x_3$~and~$x_4$.
\end{Ex}
